Thermal reactionomes reveal adaptive responses to thermal extremes in warm

### 2014-10-17 doc/ApTranscriptome_manuscript.Rmd

Thermal tolerance is a key trait governing the distributions of species and is useful in predicting their responses to climate change. To quantitatively evaluate among three hypotheses for the evolution of thermal tolerance, we describe a novel analysis of the *reactionome*; the reaction norm for all genes in an organism's transcriptome measured across an experimental gradient. We characterized the thermal reactionomes of two common ant species in the eastern U.S: the northern cool-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*. We found that at least 8% of all genes were thermally responsive. Consistent with the *enhanced defense hypothesis*, the cool-adapted *A. picea* induced expression of more genes in response to extreme temperatures than *A. carolinensis*, including heat shock proteins. In contrast, *A. carolinensis* displayed higher constitutive expression of responsive genes and required greater changes towards high, but not low, temperatures to induce changes in gene expression than *A. picea*. This pattern was consistent with the *tolerance hypothesis* at high, but not low, temperatures. We found no evidence for the *genetic assimilation hypothesis* as there was no trade-off between constitutive and inducible levels of gene expression. Collectively, our results indicate that the northern *A. picea* actively resists thermal shock by synthesizing heat shock and other protective molecules to maintain function, whereas the southern *A. carolinensis* relies upon other mechanisms to tolerate thermal stress. These results are consistent with expectations that thermally stressful habitats select for an evolutionary shift in response mechanism away from high-cost damage repair and toward lower-cost prevention.

@@

Coping with progressive changes in climate requires both short-term physiological acclimation and long-term adaptation. To better understand the relationship between these two mechanisms, we compared gene expression as a function of temperature in two closely-related ant species: the cool-climate *Aphaenogaster picea* and the warm-climate *A. carolinensis*. Surprisingly, we found divergent transcriptomic responses in these species. *Aphaenogaster picea* responded to high and low temperatures through upregulation of a larger proportion of its transcriptome with a robust heat shock response. In contrast, *A. carolinensis* required more extreme temperatures to trigger a protective response and downregulated metabolic genes. These results suggest that adaptation to chronic high temperatures involves qualitatively different mechanisms than those used to cope with low temperatures and other stressors.

@@

To address this gap of knowledge, we need information on the extent to which selection has acted upon the diversity genes and their plasticity of expression involved in thermal tolerance [@hoffmann2008; @somero2011]. In recent years, whole organism gene expression approaches (e.g. transcriptomics) using high-throughput RNA sequencing (RNAseq) technology have been widely applied to identify genes involved in thermal tolerance [@meyer2011; @teets2012; @barshis2013; @oneil2014] and other traits. Such studies typically use an ANOVA-type experimental or sampling design, with only a few environmental levels, and often find only a few dozen to hundred genes with differential expression in different thermal regimes. However, temperature and other environmental factors vary continuously in nature. As a result, such comparisons (e.g. high vs. low temperatures) are likely to miss key differences that are due not just to whether it is hot, but rather how hot it is. Continuous variation is better characterized with a reaction norm approach, which describes variation in the phenotype of a single genotype across an environmental gradient [@gomulkiewicz1992]. Reaction norms differ not only in mean values, but also in their shapes [@hoffmann2013; @murren2014], and differences in the shape of reaction norms are often greater than differences in mean values at both the species and population levels [@murren2014].

@@

```{r expression_change, echo=FALSE, warning = FALSE, fig.cap = "Matched expression type data for *A. carolinensis* (rows) and *A. picea* (columns). The color of each cell represent the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies. Blue cells are those with significant excesses, while red cells are those with significant absences."}

# plot
mh_plot <- qplot(x=Apic, y=Acar, data=md, fill=value, geom="tile", ylim = rev(levels(md$Acar))) + 
  scale_fill_gradient2(limits=c(-600, 600)) 
mh_plot
```

@@

**ADD PANEL TO FIG2 WITH DOWNREGULATION OF INTERMEDIATE TRANSCRIPTS**

```{r T_on_plot, echo=FALSE, warning = FALSE, fig.cap = "Histogram with smooth density estimate of temperature of maximum change in expression for transcripts that have increased expression at **Low** and **High** temperatures. Red bars and lines are for *A. carolinensis* while blue bars and lines are for *A. picea*."}

T_on_plot <- ggplot(Ap.T_on, aes(x = T_on, y = ..density.., colour = Species, group = Species, fill = Species)) +
    facet_grid(. ~ type) +
    geom_histogram(position = "dodge", binwidth = 3) +
    geom_density(alpha = 0.2, adjust = 3) + 
    scale_y_continuous(name="Density") +
    scale_x_continuous(name=expression("Temperature of transcript activation"))
print(T_on_plot)
```

@@

Collectively, our results indicate that these two species have distinct mechanisms of response to thermal extremes. The cool-adapated species *Aphaenogaster picea* mounts a classical cellular stress response [@kultz2005; @richter2010] with elevated gene expression at both low and high temperatures, which has been shown to impose a metabolic cost [@hoekstra2013]. In contrast, the warm-adapted *A. carolinensis* relies upon constitutive defenses [@barshis2013], and perhaps enhanced proteome stability [@ghosh2010] or behavioral responses [@sunday2014] to tolerate thermal stress. If the *enhanced defense* is hypothesis is correct, candidate genes such as heat shock proteins should be over-represented in the thermally-responsive category for *A. picea*. Of the `r nrow(hsp_all)` transcripts annotated to heat shock proteins in the transcriptome, `r length(which(hsp_responsive$A22.type %in% c("Bimodal", "High", "Low")))` were upregulated to thermal extremes in *A. picea* whereas only `r length(which(hsp_responsive$Ar.type %in% c("Bimodal", "High", "Low")))` were upregulated in *A. carolinensis*. These differences are in line with the expectations if *A. picea*, a northern species with fewer good foraging days, must forage in suboptimal conditions whereas *A. carolinensis* can, with a longer season, afford to restrain from foraging in suboptimal conditions. In fact, restrained foraging under stressful conditions by the red harvester ant *Pogonomyrmex barbatus* has been shown to increase colony fitness [@gordon2013].

Critical maximum and minimum temperatures are hypothesized to be genetically correlated [@addo-bediako2000; @hoffmann2013], but this was not evident in terms of gene expression.  Only 15-20% of transcripts upregulated in response to temp are bimodal (i.e. generalized stress response), and although upper activation thresholds differed significantly between species, lower thresholds were virtually identical. Implications ...

Given the differences in the patterns of thermal responsiveness between species (Fig. 2), it is worth noting the similarity in expression patterns between species. In both species, there are 3 -- 4 times more transcripts upregulated at low than high temperatures  (Fig. 3). A potential explanation for this pattern is that selection has been stronger on upper than lower thermal limits, thereby reducing genetic variation at high temperatures [@kingsolver2013; @kelly2013]. This explanation is consistent with the observation in *Aphaenogaster rudis* [@warren2013] and other ectotherms [@addo-bediako2000; @hoffmann2013] that critical maximum temperatures vary less among taxa than do critical minimum temperatures. Among closely-related *Drosophila* species, limits to thermal maxima appear to be due to a scarcity of genetic variation in thermal tolerance [@kellermann2012]. To the extent that the number of thermally responsive genes is related to additive genetic variation, our results also suggest that adaptability to high temperatures is more genetically limited than low temperatures, perhaps placing these northern species at risk of competitive replacement by more heat-tolerant southern species [@rodriguez-cabal2012].

A major contribution of this study is the construction of a reactionome for gene expression data (but see Hodgins-Davis et al [-@hodgins-davis2012] for a similar approach using differential expression among many levels). A number of issues have hampered RNA-seq studies to date. Namely, lists of differentially expressed transcripts are prone to false positives [@gonzalez2013], depend on the genetic background of the organism [@sarup2011], identify transcripts that are correlated, but not necessarily causative, with phenotypes [@feder2005] and are prone to "storytelling" [@pavlidis2012]. Our findings are robust to these issues as we focus on the average change in the shape of the reaction norms across many hundreds of responsive transcripts in each species. While we use gene ontology information to interpret our results, the key findings about differential plasticity of expression between species do not depend on functional annotation. Thus, our method provides an example of how RNA-seq data can be used to understand differences among species  ...

@@

>>>>

The distributions of species and their responses to climate change are in part determined by their thermal tolerances. However, little is known about how thermal tolerance evolves. To test whether evolutionary extension of thermal limits is accomplished through enhancing the cellular stress response (*enhanced defense*), consititutively elevated expression of protective genes (*genetic assimilation*) or shifting from damage resistance to passive mechanisms of thermal stability (*tolerance*), we describe a novel analysis of the *reactionome*; the reaction norm for all genes in an organism's transcriptome measured across an experimental gradient. We characterized thermal reactionomes of two common ant species in the eastern U.S the northern cool-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*. We found that at least 8% of all genes were thermally responsive. The majority of upregulation was specific to low temperatures, and consistent with the *enhanced defense hypothesis* the cool-adapted *A. picea* induced expression of more genes in response to extreme temperatures than *A. carolinensis*, including heat shock proteins. In contrast, the warm-adapted *A. carolinensis* downregulated many of the genes upregulated in *A. picea* and required more extreme temperatures to induce changes in gene expression, consistent with the *tolerance* hypothesis. We found no evidence for a trade-off between constitutive and inducible gene expression as predicted by the *genetic assimilation hypothesis*. These results are consistent with expectations that adaptive increases in upper thermal limits may require an evolutionary shift in response mechanism away from high-cost damage repair and toward lower-cost prevention.

@@

Coping with progressive changes in climate requires both short-term physiological acclimation and long-term adaptation. To better understand the relationship between these two mechanisms, we compared gene expression as a function of temperature in two closely-related ant species: the cool-climate *Aphaenogaster picea* and the warm-climate *A. carolinensis*. Surprisingly, we found divergent transcriptomic responses in these species. *Aphaenogaster picea* responded to high and low temperatures through upregulation of a larger proportion of its transcriptome with a robust heat shock response. In contrast, *A. carolinensis* required more extreme temperatures to trigger a protective response and downregulated metabolic genes. These results suggest that adaptation to chronic high temperatures involves qualitatively different mechanisms than those used to cope with rarely encountered stressors.

@@

To address this gap of knowledge, we need information on the extent to which selection has acted upon the diversity and plasticity of genes involved in thermal tolerance [@hoffmann2008; @somero2011]. In recent years, whole organism gene expression approaches (e.g. transcriptomics) using high-throughput RNA sequencing (RNAseq) technology have been widely applied to identify genes involved in thermal tolerance [@meyer2011; @teets2012; @barshis2013; @oneil2014] and other traits. Such studies typically use an ANOVA-type experimental or sampling design, with only a few environmental levels, and often find only a few dozen to hundred genes with differential expression in different thermal regimes. However, temperature and other environmental factors vary continuously in nature. As a result, such comparisons (e.g. high vs. low temperatures) are likely to miss key differences that are due not just to whether it is hot, but rather how hot it is. Continuous variation is better characterized with a reaction norm approach, which describes variation in the phenotype of a single genotype across an environmental gradient [@gomulkiewicz1992]. Reaction norms differ not only in mean values, but also in their shapes [@hoffmann2013; @murren2014], and differences in the shape of reaction norms are often greater than differences in mean values at both the species and population levels [@murren2014].

@@

@@

@@

The potential for many species to persist in face of climate change will depend in part upon their thermal tolerances. However, for most species little is known about the plasticity or adaptive changes in gene expression that underlie thermal tolerances. By using the novel *reactionome* approach, we were able to quantitatively describe changes in transcript expression across a thermal gradient. In particular, by characterizing the thermal reactionomes of the ant species *Aphaenogaster picea* and *A. carolinensis* we found patterns consistent with different mechanisms of thermal tolerance in these two species.

Under the *enhanced defense* hypothesis, stress-adapted species are hypothesized to induce a stronger and earlier response to extreme conditions. We found evidence for this hypothesis at low temperature extremes: although the lower thermal limit for *A. picea* is substantially lower than *A. carolinensis*, the two species upregulated responsive transcripts at the same threshold temperature (Fig. 2A). Moreover, the transcripts upregulated in *A. picea* included nearly the entire set upregulated in *A. carolinensis* as well as an additional set of ~1,000 transcripts (Table 1), enriched for metabolism and damage repair functions (Supplemental...). Two non-mutually exclusive hypotheses may explain this pattern. First, surviving prolonged low temperatures, such as would be experienced during overwintering, generally requires advance production of specialized cryoprotectants and a suite of preparatory physiological modifications [@denlinger2002]. The northern species *A. picea* might need to induce a greater response to survive the longer winter period. Alternatively, the response to low temperatures may reflect countergradient expression to counteract reduction in enzyme efficiency, and maintain activity as temperature declines [@lockwood2012]. This requirement may be under stronger selection in *A. picea* given the short growing season that would necessitate foraging under broader ranges of temperature (...stuble data...).

In contrast to cold tolerance, the enhanced upper thermal limit in *A. picea* is best explained by the *tolerance* hypothesis. High temperatures were associated with significantly fewer upregulated transcripts in *A. carolinensis*, and a large proportion (361 of 1,238) of those transcripts upregulated in *A. picea* were either downregulated or expressed at negligible levels overall in *A. carolinensis*. Similarly, signficantly higher temperatures were required to induce the expression response for *A. carolinensis* than *A. picea* towards the high thermal limit (Fig. 2B, D), suggesting that mechanisms other than the heat shock response are acting to maintain protein stability in face of temperature increases.

The one hypothesis that did not receive unequivocal support was the *genetic assimilation hypothesis*, which predicted that exposure to more frequent stressors is hypothesized to select for a shift from inducible to constitutive expression of stress-response transcripts. In contrast to this prediction, constitutive expression levels at 19.5°C were higher in *A. carolinensis* than *A. picea* in nearly all comparisons, with the specific notable exception of **High** transcripts in *A. picea* that were **Not Responsive** in *A. carolinensis* not because their expression was enhanced, but because they were not expressed at all. This constrasts with other recent studies on adaptation in field populations to thermal stress [@barshis2013], although in a short-term selection experiment for heat tolerance, Sikkink et al. [-@sikkink2014] also found no evidence for genetic assimilation at the expression level after 10 generations of selection for heat tolerance in *Caenorhabditis remanei*, even though there was a substantial increase in heat tolerance and loss of plasticity in the evolved populations. Both the genetic assimilation and enhanced defense routes routes to increasing thermal limits are functionally similar in that they emphasize damage prevention rather than repair. Whether a particular taxon evolves one strategy over another may be related to the intensity, frequency and duration of temperature stress in a given environment.

Collectively, our results indicate that these two species have distinct mechanisms of response to thermal extremes. The cool-adapated species *Aphaenogaster picea* mounts a classical cellular stress response [@kultz2005; @richter2010] with elevated gene expression at both low and high temperatures, which has been shown to impose a metabolic cost [@hoekstra2013]. In contrast, the warm-adapted *A. carolinensis* relies upon other constitutive defenses [@barshis2013], and perhaps enhanced proteome stability [@ghosh2010] or behavioral responses [@sunday2014] to tolerate thermal stress. Although the reactionome approach is designed to identify large-scale response patterns, this difference between species is also evident among candidate genes known to be involved in stress response. Of the `r nrow(hsp_all)` transcripts annotated to heat shock proteins in the transcriptome, `r length(which(hsp_responsive$A22.type %in% c("Bimodal", "High", "Low")))` were upregulated to thermal extremes in *A. picea* whereas only `r length(which(hsp_responsive$Ar.type %in% c("Bimodal", "High", "Low")))` were upregulated in *A. carolinensis*. These differences are in line with the expectations if *A. picea*, a northern species with fewer good foraging days, must forage in suboptimal conditions whereas *A. carolinensis* can, with a longer season, afford to restrain from foraging in suboptimal conditions. In fact, restrained foraging under stressful conditions by the red harvester ant *Pogonomyrmex barbatus* has been shown to increase colony fitness [@gordon2013].

Given the differences in the patterns of thermal responsiveness between species (Fig. 2), it is worth noting the similarity in expression patterns between species. In both species, there are 3 -- 4 times more transcripts upregulated at low than high temperatures  (Fig. 3). A potential explanation for this pattern is that selection has been stronger on upper than lower thermal limits, thereby reducing genetic variation at high temperatures [@kingsolver2013; @kelly2013]. This explanation is consistent with the observation in *Aphaenogaster rudis* [@warren2013] and other ectotherms [@addo-bediako2000; @hoffmann2013] that critical maximum temperatures vary less among taxa than do critical minimum temperatures. Among closely-related *Drosophila* species, limits to thermal maxima appear to be due to a scarcity of genetic variation in thermal tolerance [@kellermann2012]. To the extent that the number of thermally responsive genes is related to additive genetic variation, our results also suggest that adaptability to high temperatures is more genetically limited than low temperatures, perhaps placing these northern species at risk of competitive replacement by more heat-tolerant southern species [@rodriguez-cabal2012]. At the same time, critical maximum and minimum temperatures are hypothesized to be genetically correlated [@addo-bediako2000; @hoffmann2013], but this was not evident in terms of gene expression.  Only 15-20% of transcripts upregulated in response to temperature are bimodal (i.e. generalized stress response), and although upper activation thresholds differed significantly between species, lower thresholds were virtually identical (Fig. 2). This suggests that species do not face a fundamental trade-off between these two limits and may be able to shift upper and lower thermal limits independently to match requirements of more seasonally variable environments.

A major contribution of this study is the construction of a reactionome for gene expression data (but see Hodgins-Davis et al [-@hodgins-davis2012] for a similar approach using differential expression among many levels). A number of issues have hampered RNA-seq studies to date. Namely, lists of differentially expressed transcripts are prone to false positives [@gonzalez2013], depend on the genetic background of the organism [@sarup2011], identify transcripts that are correlated, but not necessarily causative, with phenotypes [@feder2005] and are prone to "storytelling" [@pavlidis2012]. Our findings are robust to these issues as we focus on the average change in the shape of the reaction norms across many hundreds of responsive transcripts in each species. While we use gene ontology information to interpret our results, the key findings about differential plasticity of expression between species do not depend on functional annotation. Thus, our method provides an example of how RNA-seq data can be used to understand gain meaningful insight on the process of evolution.

@@

Figure 1. Matched expression type data for *A. carolinensis* (rows) and *A. picea* (columns). The color of each cell represent the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies. Blue cells are those with significant excesses, while red cells are those with significant absences.

Figure 2. Histogram with smooth density estimate of temperature of maximum change in expression for transcripts that have increased expression at **Low** and **High** temperatures. Red bars and lines are for *A. carolinensis* while blue bars and lines are for *A. picea*.

Figure 3. Scatterplots of (A) constitutive expression in *A. picea* against degree of induction for upregulated transcripts in *A. carolinensis* and (B) constitutive expression in *A. carolinensis* against degree of induction for upregulated transcripts in *A. picea.*. For each species, the transcripts are split into expression category based on the ... The top row is for *High* transcripts, middle is for *Intermediate* transcripts and bottom row is for *Low* transcripts.

# Figures

![Figure 1](matched_observations_Fig1.tiff)

![Figure 2](crit_temp_regulation_Fig2.tiff)

![Figure 3](constitutive_vs_upreg_Fig3.tiff)
